---
title: "Reproducible Research: Peer Assessment 1"
#Michael Nichols
output: 
  html_document:
    keep_md: true
---
###Basic assumptions:
- assumes activity data downloaded and unzipped into working directory

Load packages
```{r load packages}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(knitr)
```


## Loading and preprocessing the data

Loading and preprocessing the data (with initial assumption that we can ignore missing values)
```{r Loading and preprocessing}
file <- "activity.csv"
activity <- read.csv(file)
activity <- subset(activity, is.na(activity$steps) == FALSE)
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day
2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day

Find mean total number of steps taken per day
```{r aggregate for mean}
DailySteps <- aggregate(activity$steps, by = list(activity$date), sum)
names(DailySteps) <- c("Day", "Steps")
```

create histogram of daily steps
```{r histogram}
p <- ggplot(DailySteps, aes(x = Steps)) + geom_histogram(color = "black", fill = "grey", binwidth = 1000) + ggtitle("Total Number of Steps Per Day")
```

annotate histogram with mean and median data and then display
```{r histogram2}
p <- p + geom_vline(aes(xintercept = mean(Steps)), color = "red", linetype = "dashed", size = 1)
p <- p + geom_vline(aes(xintercept = median(Steps)), color = "green", linetype = "dotted", size = 1) 
p <- p + labs(y = "Frequency (in days)") 
p <- p + ggtitle(labs(subtitle = "Dashed Red Line ~ Average & Dotted Green Line ~ Median"))
print(p)
```

Calculate mean and median steps per day
```{r stats}
meanSteps <- mean(DailySteps$Steps)
print(meanSteps)
medianSteps <- median(DailySteps$Steps)
print(medianSteps)
```


## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

Aggregate steps by 5-minute interval (x-axis) by the average number of steps taken, averaged across all days (y-axis)
```{r stats aggregates by interval}
StepsPer5 <- aggregate(activity$steps, by = list(activity$interval), mean)
names(StepsPer5) <- c("Interval", "AverageSteps")
```

Create time series plot and then display
```{r stats time series plot}
p2 <- ggplot(NULL, aes(y = StepsPer5$AverageSteps, x = StepsPer5$Interval)) 
p2 <- p2 + geom_line() + geom_point() 
p2 <- p2 + labs(y = "Average Steps per 5 Minute Interval", x = "5 Minute Time Interval Number") 
p2 <- p2 + ggtitle(label = "Time Series Plot of Average Steps Per 5-Min Time Interval", subtitle = "Time Intervals: 0 = '12:00-12:05AM' to 2355 = '11:55PM - 12:00AM'")
print(p2)
```

Display highest average step count per interval & interval in which the highest average occurs
```{r stats highest interval}
Max5MinSteps <- max(StepsPer5$AverageSteps)
print(Max5MinSteps)

max5MinInterval <- StepsPer5[StepsPer5$AverageSteps == Max5MinSteps,]
print(max5MinInterval)
```


## Imputing missing values

Reread in dataset & count number of missing values & percentage of total
```{r stats reload original data}
activity2 <- read.csv(file)

percentMissing <- sum(is.na(activity2$steps)) / (sum(!is.na(activity2$steps)) + sum(is.na(activity2$steps)))
print(percentMissing)
```

Replace missing vales with average per 5 min interval (StepsPer5 data frame)

```{r simulate NAs}
for (i in 1:length(activity2$steps)) {
  if (is.na(activity2$steps[i] == TRUE)) {
  activity2$steps[i] <- StepsPer5$AverageSteps[match(activity2$interval[i], StepsPer5$Interval)]
  }
}
```

Show new dataset
```{r display new dataset without NAs}
str(activity2)
summary(activity2$steps)
```

## Are there differences in activity patterns between weekdays and weekends?

### 4. Make a histogram of the total number of steps taken each day and Calculate and report the  mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Create new data frame of total daily steps with average data for missing values
```{r aggregated full data frame}
DailySteps3 <- aggregate(activity2$steps, by = list(activity2$date), sum)
names(DailySteps3) <- c("Day", "FullSteps")
```

Create histogram of daily steps with substituted missing values
```{r full histogram}
p3 <- ggplot(DailySteps3, aes(x = FullSteps)) + geom_histogram(color = "black", fill = "grey", binwidth = 1000) + ggtitle(label = "Total Number of Steps Per Day")

# annotate histogram with mean and median data
p3 <- p3 + geom_vline(aes(xintercept = mean(FullSteps)), color = "red", linetype = "dashed", size = 1) 
p3 <- p3 + geom_vline(aes(xintercept = median(FullSteps)), color = "green", linetype = "dotted", size = 1) 
p3 <- p3 + labs(y = "Frequency (in days)", x = "Total Steps") 
# add subtitle explaining mean and median lines
p3 <- p3 + ggtitle(labs(subtitle = "Simulated Data for Missing Values; Dashed Red Line ~ Average & Dotted Green Line ~ Median"))

#print histogram
print(p3)
```

Calculate mean and median
```{r full mean and median}
meanFullSteps <- mean(DailySteps3$FullSteps)
print(meanFullSteps)
medianFullSteps <- median(DailySteps3$FullSteps)
print(medianFullSteps)
```

Compare daily step calculations with missing vs simulated data
### Summary below shows the substitutions by interval worked very well, keeping the mean & median the same; the major change only comes in the sum of steps, which rises by 86000 (8 avg days)
```{r compare summaries}
DailyComp <- merge(DailySteps, DailySteps3, all.y = TRUE)
summary(DailyComp)
diffSteps <- sum(DailySteps3$FullSteps) - sum(DailySteps$Steps)
print(diffSteps)
```

### Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r weekday / weekend factor}
#reformat date variable as date (from factor)
activity2$date <- as.Date(as.character(activity2$date))

#append weekday var to data frame
activity2 <- mutate(activity2, Weekday = weekdays(activity2$date))

#adjust weekday variable (as factor) for weekday vs weekend
for (i in 1:length(activity2$steps)) {
  if (activity2$Weekday[i] %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")) {
    activity2$Weekday[i] <- "Weekday"
  }  else {
    activity2$Weekday[i] <- "Weekend"
  }
}
activity2$Weekday <- as.factor(activity2$Weekday)
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was created using simulated data:

```{r subsets by weekday / weekend and new averages}
# create 2 subsets and aggregate by average per interval once again 
weekdaySteps <- subset(activity2, activity2$Weekday == "Weekday")
weekendSteps <- subset(activity2, activity2$Weekday == "Weekend")

StepsPer5Weekday <- aggregate(weekdaySteps$steps, by = list(weekdaySteps$interval), mean)
StepsPer5Weekend <- aggregate(weekendSteps$steps, by = list(weekendSteps$interval), mean)
names(StepsPer5Weekday) <- c("Interval", "AverageWeekdaySteps")
names(StepsPer5Weekend) <- c("Interval", "AverageWeekendSteps")

#new data frame for 5 min intervals by weekday & weekend
StepsPer5Grouped <- merge(StepsPer5Weekday, StepsPer5Weekend)
```

```{r time series plots by weekday vs weekend}
# create stacked ggplots with labels on only selected charts and consistent y axis limits
p4 <- ggplot(NULL, aes(y = StepsPer5Grouped$AverageWeekdaySteps, x = StepsPer5Grouped$Interval)) 
p4 <- p4 + geom_line() + geom_point() + ylim(0, 235)
p4 <- p4 + ggtitle(labs(y = "Average Number of Steps", x = ""))
p4 <- p4 + ggtitle(label = "Time Series Plot of Average Steps Per 5-Min Time Interval", subtitle = "Weekday")


p5 <- ggplot(NULL, aes(y = StepsPer5Grouped$AverageWeekendSteps, x = StepsPer5Grouped$Interval)) 
p5 <- p5 + geom_line(color = "grey") + geom_point(color = "purple") + ylim(0, 235) 
p5 <- p5 + ggtitle(labs(y = "", x = "5 Minute Time Interval")) 
p5 <- p5 + ggtitle(labs(subtitle = "Weekend"))

grid.arrange(p4, p5, ncol = 1)
```


